Dynamic strategies for yaw and induction control of wind farms based on large-eddy simulation and optimization

### 2018-01-07 reply_to_reviewer1.tex

\paragraph{Reviewer} \textit{The present article investigates the improvement on wind-farm power by performing different wind turbine optimization strategies (yaw and induction) modeled using large-eddy simulations (LESs) combined with an adjoint-based optimization method. Two test cases are considered, one characterized by uniform incoming flow conditions (turbulence void), and another where a precursor LES is used to seed inflow turbulence in the domain containing the 4 x 4 wind turbines layout. The authors present new results about combining axial induction and dynamic yaw control. Although for the relatively small wind farm under consideration, the yaw control over-performs induction control, the authors present interesting analysis on the potential of combining the two methods. The manuscript is overall well written and organized, and the majority of the results are correctly interpreted and clearly articulated. Therefore, I recommend publication after minor 	revisions are made. Here below a list a series of concerns/clarifications that need to be addressed prior to its publication.}

\section*{Comments}
\hrulefill

\paragraph{1. Reviewer} \textit{The authors use an actuator disk model that does not account for rotational effects (line 158). However, in lines 362-364 the authors state that their LES model output exhibits curling of the wakes as a result of the control strategy. What is the impact of the nonrotating actuator disk approach on wake curling and turbine control? It would be interesting to repeat on of the cases, in particular I3Y, including the tangential forces on the turbine, and report on the differences in the dynamics as well as power extraction and wind farm efficiency found.}

\paragraph{Response} We thank the reviewer for this comment. Indeed, a non-rotating actuator disk model (ADM) is used in the current study. In a recent paper by Howland \emph{et al.} [24], the wake structure behind turbines in yaw was studied using both experimental and numerical techniques. It was found that both a non-rotating ADM and a rotating actuator line model (ALM) produce similar curled wakes, and that the wake center was deflected to a similar degree in both models. The main dissimilarity in the wakes from both models is a rotation-induced top-down asymmetry in the ALM case. Therefore, it is expected that an ADM with tangential forces would yield similar results: a comparable wake deflection with rotation-induced top-down asymmetry. Therefore, the prevailing control mechanisms identified in the current paper are expected to also be found in more detailed turbine models, albeit with slight differences in e.g. yaw angles of the upstream turbines.

We agree that performing optimal yaw control simulations with more detailed turbine models would yield interesting results for comparison with the current ADM. This is subject of parallel ongoing work.

We have included the explanation above in the revised manuscript as follows:
\begin{itemize}
	\item Section 4.2.2 (Static yaw regime - wake redirection, \red{line XX, p YY}): \\``... redirecting the wakes sufficiently. \revision{It is important to note that such wake redirection and curling is not simply an artifact of the currently used non-rotating ADM, since similar wake characteristics were also observed in rotating actuator line models [24].} The static yaw wake ...''
	\item Section 5 (Conclusions and future work, \red{line XX, p YY}): \\``\revision{The current optimization studies have been performed using relatively simple non-rotating actuator disk models for the wind turbine rotors without the inclusion of towers or nacelles. Although it has been shown that, by using this approach, the fidelity of far wake characteristics is satisfactory, the credibility of the current power gains and control strategies in practice could be further increased through wind-tunnel testing and similar optimization studies with more detailed turbine models, e.g. by adding drag forces for the tower and nacelle or by representing the rotor with an actuator line model. These are left for future work.} Finally, note that the current study optimizes power at all costs, and did not account for any undesirable loading characteristics. These should also be considered upon thoroughly comparing induction and yaw control for practical purposes. \revision{In this latter case, a detailed turbine representation is of high importance.}''
\end{itemize}

[24] M. F. Howland, J. Bossuyt, L. A. Martinez-Tossas, J. Meyers, C. Meneveau, ``Wake Structure of Wind Turbines in Yaw under Uniform Inflow Conditions'' Journal of Renewable and Sustainable Energy 8, 043301 (2016)

@@

\hrulefill

\paragraph{2. Reviewer} \textit{Figure 12a displays a strong power recovery for rows 3 and 4 for the cases where yaw control is applied (this effect is less evident for the induction control cases), which results in a wind farm efficiency increase of ~20\% with respect to the reference case. I wonder how much of that improvement is due to the reduced number of wind turbine rows (only 4), and the ability to quickly adjust to a much higher power extraction level as the end of the wind farm is approached. It would be desirable to perform at least one more simulation (I3Y) with an extended domain and wind farm, doubling the number of turbine rows from 4 to 8. Such case will reveal the trend of the increase in power extraction and wind farm efficiency as the size of the wind farm is enlarged, and will help extrapolate to what the gain would be for a more realistic wind farm size. In addition, the authors could consider another case where the use of periodic boundary conditions in the streamwise direction, as an asymptotic limit for an “infinite” wind farm, allowing to keep the size of the simulation (although not informative of row dependency within the wind farm). These two additional cases will further increase the robustness of the analysis and conclusions regarding the combined induction/yaw control approach proposed, and the finite wind farm effects.}

\paragraph{Response} This is an important comment. We agree that the current layout is very suited for wake redirection through yaw control. More specifically, the relatively short streamwise extent of the farm and the aligned layout are favorable for wake redirection since wakes do not interact with downstream turbines in other columns.

The desired additional I3Y case suggested by the reviewer is subject of current research and has been submitted in a conference paper at the Science of Making Torque conference in June. In this work, we will perform induction, yaw, and combined control (equivalent to I3, Y, and I3Y) for wind farms with 12 turbine rows in both aligned and staggered conditions such that we can draw properly funded conclusions for larger wind farms with various layouts. The computational time of these simulations however disallow them to be included in the current paper.

In the submitted version of the manuscript, we already indicated in the summary that the quantified relative profitability of induction versus yaw control have to be interpreted with the given wind-farm layout in mind.

In the revised version of the manuscript, we have made this remark more explicit by mentioning this every time the power gains of yaw control are compared to induction control, i.e.: 
\begin{itemize}
	 \item Abstract (\red{line XX, p 1}): \\ ``... We study a 4 $\times$ 4 aligned wind farm, and find that \revision{for this farm layout} yaw control is more effective than induction control. ...''
	 \item Section 4.2.1 (Uniform inflow power extraction, \red{line XX, p YY}): \\``... close to freestream conditions. \revision{Remark that the favorability of yaw over induction control is quantified here for a relatively small set of aligned, and care should be taken when extrapolating these results to wind farms of larger size and/or different layouts (see further discussion in Section 5).}''
	 \item Section 4.4 (Discussion, \red{line XX, p YY}): \\``... for the current setup \revision{with a relatively small aligned wind farm}, all yaw-enabled cases consistently outperform cases based solely on axial induction control. ...''
	 \item Section 5 (Conclusions and future work, \red{line XX, p YY}): \\``...depending on their position within the farm. \revision{Moreover, also the wind-farm layout is expected to affect which control mechanism yields more power gains, e.g. the opportunities for wake redirection in aligned wind farms are much better than in staggered wind farms, where deflected wakes will more easily hit turbines in neighboring columns. The question of which control strategy is more effective in larger wind farms with various layouts is subject of current investigation.} ''
\end{itemize}

The suggested infinite wind farm optimal control case as suggested by the reviewer has been performed for induction control in Goit \& Meyers 2015, J Fluid Mech 768. Comparing optimal control in periodic farms to finite farms can be problematic: experience has shown that the optimizer tends to increase power by slowing down the overall boundary layer in a transient manner and in infinite farms, requiring an ad-hoc penalty term to be used by Goit \& Meyers. This large-scale slow-down is not possible in finite wind-farms, where the inlet is prescribed by the precursor simulation. \red{The entire idea of using power gains from infinite farms to draw conclusions for finite farms is hence questionable, and warrants its own further investigation which is out of scope of the current work.}

@@

\hrulefill

\paragraph{3. Reviewer} \textit{Line 226-230. How is that time related to an eddy and advection scales? What are the implications going to real large-scale forcing variability? Please comment on that.}

\paragraph{Response} Regarding the relation of the user-defined time horizons we have related them to the characteristic timescale of the wind farm, i.e. the wind-farm flowthrough time as follows (\red{line XX, p YY}):

``
... total control time of 900~s.  \revision{Note that, given a wind-farm length of $3 \times 6D = 1 800$ m and a free-stream velocity $U_\infty = 8$ m s$^{-1}$, the wind-farm flowthrough time is approximately 225 s. Hence, the total control time comprises about 4 flowthroughs, and the optimization horizon within each window covers about 1.33 flowthrough times. Because of the latter, each turbine row has the opportunity to take into account interaction with every downstream row, and the sensitivity of power gains to the optimization horizon $T$ is expected to be limited.} All time-averaged results ...
''

Regarding the implications of real large-scale forcing variability, the same time horizons could be used in principle. However, quantifying potential power gains for the different in an actual wind-farm site would obviously require to take annual variability in wind speed and direction into account. We added a comment on this in the Summary section (\red{line XX, p YY}):

`` 
... atmospheric conditions for particular wind farms. \revision{Also, throughout this work we considered a constant wind direction and flow forcing. When quantifying potential power gains for real wind farms, the annual distribution of wind speed and velocities at the given site should be taken into account in the flow forcing.} ...
''

\red{Weet niet of eddy time scales hier echt relevant zijn...}

@@

\hrulefill

\paragraph{4. Reviewer} \textit{Line 291-292. It would be beneficial to include the turbulent case also, making it a 4-panel figure. That case is more representative of an ABL flow and therefore provides the reader a better sense of the optimization convergence in a more realistic scenario.}

\paragraph{Response} As indicated in the text, the turbulent inflow case follows a similar behavior as the uniform inflow. Including both uniform and turbulent convergence characteristics would in our opinion draw the attention away from the more important results in the paper. As a compromise, we replaced the uniform optimization convergence with the turbulent one in the revised manuscript (also shown below).

\begin{figure}[h]
	\includegraphics[width=\textwidth]{figure4_revised}
	\caption{Convergence behavior for the second optimization window in \revision{turbulent} inflow. \emph{a) }: Cost functional $\mathscr{J}$. \emph{b) }: PDE evaluations (forward or adjoint). Markers represent every fifteenth iteration step.}
\end{figure}

@@

\hrulefill

\paragraph{5. Reviewer} \textit{Line 382-385. What about the peak that shows up in all case for $St < 0.1$ ($\approx$ 0.05)? Please explain.}

\paragraph{Response} This peak, occuring at the first non-zero frequency, is an artifact of the way the spectral density estimates are computed. The data is detrended by removing the mean before computing the PSD estimate using Welch's method, which divides the data in overlapping windows for increased statistical convergence. This can cause loss of frequency resolution in the lowest frequencies of the spectrum resulting in energy bleeding into other frequencies including the zero frequency which represents the mean component (which should be zero in theory due to the removal of the mean). We fixed this by simply not detrending the data prior to computing the PSD estimate, resulting in a smooth behavior at the lowest frequencies with no spuriously apparent peaks.

The large spectral content at low frequencies points towards long-time trends in the yaw angles, e.g. for turbines transitioning from the meandering regime $(\overline{\theta} = 0^{\circ})$ to the redirection regime $(\overline{\theta} = 30^{\circ})$. The revised versions of figures 9 and 15 are shown below. \red{WM: Denk niet dat we hier expliciet in de tekst naar moeten verwijzen? Lijkt me vrij self-explanatory..}

\begin{figure}
	\includegraphics[width=\textwidth]{figure9_revised}
	\caption{Revised version of Figure 9. See revised manuscript.}
\end{figure}

\begin{figure}
	\includegraphics[width=\textwidth]{figure15_revised}
	\caption{Revised version of Figure 15. See revised manuscript.}
\end{figure}

### 2018-01-07 MuntersMeyers_DynamicStrategiesControl_revision_WM01.tex
Update everything for first draft revisions

% Abstract (Do not use inserted blank lines, i.e. \\) 
\abstract{In wind farms, wakes originating from upstream turbines cause reduced energy extraction and increased loading variability in downstream rows. The prospect of mitigating these detrimental effects through coordinated controllers at the wind-farm level has fueled a multitude of research efforts in wind-farm control. The main strategies in wind-farm control are to influence the velocity deficits in the wake by deviating from locally optimal axial induction setpoints on the one hand, and steering wakes away from downstream rows through yaw misalignment on the other hand. The current work investigates dynamic induction and yaw control of individual turbines for wind-farm power maximization in large-eddy simulations. To this end, receding-horizon optimal control techniques combined with continuous adjoint gradient evaluations are used. We study a 4 $\times$ 4 aligned wind farm, and find that yaw control is more effective than induction control, both for uniform and turbulent inflow conditions. Analysis of optimal yaw controls leads to the definition of two simplified yaw control strategies, in which wake meandering and wake redirection are exploited respectively. Furthermore it is found that dynamic yawing provides significant benefits over static yaw control in turbulent flow environments, whereas this is not the case for uniform inflow. Finally, the potential of combining overinductive axial induction control with yaw control is shown, with power gains that approximate the sum of those achieved by each control strategy separately.}

@@

In contrast, the current manuscript follows the approach first introduced by Goit and Meyers \cite{goit2015optimal}: instead of focusing on developing a practically realizable wind-farm controller, wind-farm performance is optimized directly in an accurate model with full state information, i.e. using large-eddy simulations of the filtered Navier--Stokes equations (see Figure \ref{fig:block_diag}b). In this way, instead of being parametrized, the turbulent interactions between wind turbines and the atmospheric boundary layer flow are explicitly resolved using adequate spatio-temporal resolution. In this way, the wind turbines can be used as flow actuators: by dynamically controlling individual thrust setpoints and yaw rates of the turbines we aim to optimally influence the ABL flow such that aggregate wind-farm power extraction is maximized.  Note that this approach is unfeasible for practical control, both due to excessive computational cost and the incomplete state information in practice. However, it allows to benchmark the potential of increasing energy through wind-farm control, and can help identify control mechanisms to be harnessed by practical controllers. Also here, a receding-horizon methodology is used. Within each control window, we consider the following PDE-constrained optimal control problem, in which we maximize the aggregate wind-farm power over a time horizon $T$:

@@

The wind turbines are modeled using a standard non-rotating actuator disk model (see, e.g., Ref \cite{meyers2010large}). For each turbine $i (= 1 \dots N_t)$, two state variables are defined. First, the turbine thrust coefficient $\ctihat$ is obtained by applying a one-sided exponential time-filter with time constant $\tau$ to the thrust coefficient control setpoint $\cti$ (Equation \ref{eq:ctihat_constraint}). Second, the turbine yaw angle $\theta_i$ is defined in a straightforward way using the yaw rate control $\omega_i$ (Equation \ref{eq:omega_constraint}). Subsequently, the thrust force enacted by the turbine upon the flow can be parametrized as $\bs{f}_{i}(\bs{x}) = - (1/2) \ctihat V_i^2 \R_i (\bs{x}) \eperpi$. Here, $\R_i(\bs{x}) = \sint G(\bs{s} - \bs{x})~H\big[D/2 - \vert\vert \bs{s} - \bs{t}_i \vert\vert_2\big]~\diracdelta \big[ (\bs{s} - \bs{t}_i)\cdot \eperpi \big] \ds$ is a smoothed geometric footprint of the rotor with diameter $D$ at location $\bs{t}_i$ on the LES grid, using a Gaussian kernel $G$ with characteristic filter width $\Delta$. Further, $V_i = (1/A_i) \int_{\Omega} \R_i(\bs{x}) \utilde \cdot \eperpi \dx$ is the axial velocity averaged over the rotor disk with area $A_i$, and the rotor-perpendicular unit vector $\eperpi = \ex \cos \theta_i + \ey \sin \theta_i$ defines the orientation of the rotor disk using the yaw angle state $\theta_i$. Analogously, a rotor-parallel unit vector $\etransi = - \bs{e}_x \sin \theta_i + \bs{e}_y \cos \theta_i$ is defined, which will be used further below. The power extraction of the turbine is calculated as $P_i = (1/2) C_{P,i}' V_i^3 A_i$, with power coefficient $C_{P,i}' = a \ctihat$. The coefficient in the latter linear relation is set to $a = 0.9$ to mitigate overprediction of turbine power on present-day simulation grids as discussed in Ref. \cite{munters2017optimal}. Both the thrust coefficient setpoint controls $\cti$ and the yaw rate controls $\omega_i$ are subjected to technical box constraints in Equations \eqref{eq:boxct_constraint} - \eqref{eq:boxomega_constraint}. As discussed in Section \ref{sec:case_definition} and Table \ref{tab:case_definition}, the control cases considered in this manuscript are differentiated based on the parameter values of these box constraints. The system states are grouped in the vector $\bs{q} = [\utilde(\bs{x},t),~\ptilde(\bs{x},t),~\ctnhat{1}(t), \dots, \ctnhat{N_t}(t),~ \theta_1(t), \dots, \theta_{N_t}(t)] = [\utilde(\bs{x},t),~\ptilde(\bs{x},t),~\widehat{\bs{C}}_T'(t) ,~ \bs{\theta}(t)]$ and the controls, consisting of the individual unsteady thrust coefficient setpoints and yaw rates of every turbine, are denoted as $\bs{\varphi} = [\bs{C}_T'(t),~\bs{\omega}(t)]$.%[\ctn{1},\dots,\ctn{N_t},~\omega_1,\dots,\omega_{N_t}]$.

@@

\subsection{Simulation setup}
\noindent The simulation domain has a \revision{total dimension of $L_x \times L_y \times L_z = 4.8 \times 2.4 \times 1 $ km$^3$, with $L_x, L_y$, and $L_z$ the streamwise, transversal, and vertical extent of the domain respectively. This domain is discretized on a grid of $N_x \times N_y \times N_z = 256 \times 128 \times 128$ gridpoints.} The final $10\%$ of the streamwise extent of the domain is used as a fringe region to impose inflow conditions. All cases apply periodic boundary conditions on transversal domain boundaries.

Two different sets of simulations are considered. The first set is subjected to a \revision{constant} and uniform inflow with $U_\infty = 8$ m s$^{-1}$. In this case, the top and bottom boundaries are treated with a symmetry condition, \revision{and the vertical location of the turbine hubs $z_h$ is placed at the vertical mid-plane of the domain, i.e. $z_h/L_z = 0.5$,} in order to minimize any influence of vertical domain boundaries. The second set applies unsteady turbulent inflow conditions, which are generated in a pressure-driven precursor simulation of an unperturbed turbulent boundary layer with shifted periodic boundary conditions \cite{munters2016shifted}, driven by a pressure gradient of $\partial_x p_\infty/\rho = 2.5 \times 10^{-4}$ m s$^{-2}$, resulting in a freestream wind speed at hub height slightly higher than 8 m s$^{-1}$. In this second case, the top boundary is treated with a stress-free condition, and a high-Reynolds number wall model is used at the bottom. The turbines are placed at a height of $z_h/L_z = 0.1$ from the wall. The wall model adds a wall stress consistent with a rough-wall logarithmic boundary layer profile with roughness length $z_0/L_z = 10^{-4}$, resulting in a inflow turbulence intensity of about 8\% at hub height.

Within the context of the receding-horizon methodology introduced in Section~\ref{sec:meth}, the optimization is performed over 6 time windows. Each window applies an optimization horizon $T=300$~s and a flow advancement time of $T_A = T/2= 150$~s, resulting in a total control time of 900~s. All time-averaged results shown below are based on statistics gathered after 300~s, so that wake propagation time lag at startup is excluded. To limit computational efforts, we perform a fixed amount of 80 L--BFGS--B iterations in each window, corresponding to approximately 160 PDE simulations (forward or adjoint). Although formal convergence is not achieved for any of the cases after this amount of iterations, it will be shown later that the relative ordering of the cases based on power extraction is clear. Computational costs amount to approximately 35~000 corehours per optimization case for all 6 time windows combined on 112 Intel Broadwell architecture cores connected by an EDR Infiniband network. The initial guess for the optimal controls within each window is a steady thrust coefficient setpoint $C_T'=2$  and yaw rate $\omega = 0^\circ$ s$^{-1}$ for all turbines. Simulation and optimization parameters are summarized below in Table~\ref{tab:case_definition}.

Simulations are initialized as follows: for the turbulent inflow case, a fully-developed high-Reynolds number turbulent boundary layer is generated in the precursor inflow simulation by simulating a perturbed logarithmic flow until a statistically-stationary state is achieved. For both the turbulent and the uniform inflow cases the main wind-farm simulation is then advanced in time with the proper inflow conditions for approximately five wind-farm flow-through times, after which the effects of wind-farm start-up transients have subsided. These statistically stationary wind-farm flows then serve as initial conditions for the control cases defined below. Planviews at hub height of these initial conditions are shown in Figure~\ref{fig:initial_conditions_flow_yaw}. It can be seen from the Figure~\ref{fig:initial_conditions_flow_yaw}a that, given laminar inflow conditions, turbine wakes are very deep and remain stable up until row 3, after which meandering and vortex shedding starts to occur. These flow conditions can be regarded as somewhat artificial since, in the atmospheric boundary layer, background turbulence typically causes instant wake transition behind the turbine rotor. Nevertheless, this setup allows us to investigate control dynamics more easily, as we can rule out any reaction of the optimization to the background variability. Furthermore, suppressed wake meandering and reduced turbulence levels have been known to occur in wind farms submerged in stably stratified atmospheric boundary layers \cite{larsen2009dependence, machefaux2016experimental}, hence rendering the current case somewhat relevant for such situations. Figure~\ref{fig:initial_conditions_flow_yaw}b shows that, for the turbulent inflow case, the background turbulence causes wake instabilities instantly, leading to enhanced wake recovery and much more complex unsteady wake interactions between turbines.

\begin{figure}
	\includegraphics[width=\textwidth]{figure3}
	\caption{Planview at hub height $z_h$ of initial conditions for optimal control cases. \emph{a)} Uniform and \revision{constant} inflow. \emph{b)} Turbulent inflow. Colors represent instantaneous streamwise velocity $\widetilde{u}_x$ in m s$^{-1}$. The black lines represent wind turbine locations. The black dashed line indicates the start of the fringe region. R1 -- R4 and C1 -- C4 indicate the numbering convention used for turbine rows and columns respectively. \label{fig:initial_conditions_flow_yaw}}
\end{figure}

@@

Similar to earlier studies \cite{munters2016turbulent, munters2016effect}, also two induction control cases are considered, in which turbines can adapt their thrust coefficient $C_T'$ with a characteristic timescale $\tau = 5$ s: an underinductive case where turbines are restricted to lowering $C_T'$ below the Betz-optimal value if desired (I2, with $\ctmax = 2$), and an overinductive case in which also higher $C_T'$ values resulting in increased thrust forces are allowed (I3, with $\ctmax = 3$). Finally, for both the latter cases, also the option of combining dynamic induction control with the yaw control of case Y is considered, defining the yaw--induction cases I2Y and I3Y respectively. The cases investigated in this study are summarized in Table~\ref{tab:case_definition}.

@@

Further results show that, for the current setup, all yaw-enabled cases consistently outperform cases based solely on axial induction control. However, the gap between both is significantly larger for uniform inflow than for turbulent inflow. The main reason for this is that the yawing cases perform better in the absence of inflow turbulence due to the fact that a \revision{constant incoming} flow allows wakes to be redirected more easily.

>>>>

% Abstract (Do not use inserted blank lines, i.e. \\) 
\abstract{In wind farms, wakes originating from upstream turbines cause reduced energy extraction and increased loading variability in downstream rows. The prospect of mitigating these detrimental effects through coordinated controllers at the wind-farm level has fueled a multitude of research efforts in wind-farm control. The main strategies in wind-farm control are to influence the velocity deficits in the wake by deviating from locally optimal axial induction setpoints on the one hand, and steering wakes away from downstream rows through yaw misalignment on the other hand. The current work investigates dynamic induction and yaw control of individual turbines for wind-farm power maximization in large-eddy simulations. To this end, receding-horizon optimal control techniques combined with continuous adjoint gradient evaluations are used. We study a 4 $\times$ 4 aligned wind farm, and find that \revision{for this farm layout} yaw control is more effective than induction control, both for uniform and turbulent inflow conditions. Analysis of optimal yaw controls leads to the definition of two simplified yaw control strategies, in which wake meandering and wake redirection are exploited respectively. Furthermore it is found that dynamic yawing provides significant benefits over static yaw control in turbulent flow environments, whereas this is not the case for uniform inflow. Finally, the potential of combining overinductive axial induction control with yaw control is shown, with power gains that approximate the sum of those achieved by each control strategy separately.}

@@

In contrast, the current manuscript follows the approach first introduced by Goit and Meyers \cite{goit2015optimal}: instead of focusing on developing a practically realizable wind-farm controller, wind-farm performance is optimized directly in an accurate model with full state information (see, e.g. Refs. \cite{goit2015optimal,goit2016optimal,munters2016effect,munters2017optimal}), i.e. using large-eddy simulations of the filtered Navier--Stokes equations (see Figure \ref{fig:block_diag}b). In this way, instead of being parametrized, the turbulent interactions between wind turbines and the atmospheric boundary layer flow are explicitly resolved using adequate spatio-temporal resolution. In this way, the wind turbines can be used as flow actuators: by dynamically controlling individual thrust setpoints and yaw rates of the turbines we aim to optimally influence the ABL flow such that aggregate wind-farm power extraction is maximized.  Note that this approach is unfeasible for practical control, both due to excessive computational cost and the incomplete state information in practice. However, it allows to benchmark the potential of increasing energy through wind-farm control, and can help identify control mechanisms to be harnessed by practical controllers. Also here, a receding-horizon methodology is used. Within each control window, we consider the following PDE-constrained optimal control problem, in which we maximize the aggregate wind-farm power over a time horizon $T$:

@@

The wind turbine rotors are modeled using a standard non-rotating actuator disk model (see, e.g., Ref \cite{meyers2010large}). \revision{Turbine towers are not included.} For each turbine $i (= 1 \dots N_t)$, two state variables are defined. First, the turbine thrust coefficient $\ctihat$ is obtained by applying a one-sided exponential time-filter with time constant $\tau$ to the thrust coefficient control setpoint $\cti$ (Equation \ref{eq:ctihat_constraint}). Second, the turbine yaw angle $\theta_i$ is defined in a straightforward way using the yaw rate control $\omega_i$ (Equation \ref{eq:omega_constraint}). Subsequently, the thrust force enacted by the turbine upon the flow can be parametrized as $\bs{f}_{i}(\bs{x}) = - (1/2) \ctihat V_i^2 \R_i (\bs{x}) \eperpi$. Here, $\R_i(\bs{x}) = \sint G(\bs{s} - \bs{x})~H\big[D/2 - \vert\vert \bs{s} - \bs{t}_i \vert\vert_2\big]~\diracdelta \big[ (\bs{s} - \bs{t}_i)\cdot \eperpi \big] \ds$ is a smoothed geometric footprint of the rotor with diameter $D$ at location $\bs{t}_i$ on the LES grid, using a Gaussian kernel $G$ with characteristic filter width $\Delta$. Further, $V_i = (1/A_i) \int_{\Omega} \R_i(\bs{x}) \utilde \cdot \eperpi \dx$ is the axial velocity averaged over the rotor disk with area $A_i$, and the rotor-perpendicular unit vector $\eperpi = \ex \cos \theta_i + \ey \sin \theta_i$ defines the orientation of the rotor disk using the yaw angle state $\theta_i$. Analogously, a rotor-parallel unit vector $\etransi = - \bs{e}_x \sin \theta_i + \bs{e}_y \cos \theta_i$ is defined, which will be used further below. The power extraction of the turbine is calculated as $P_i = (1/2) C_{P,i}' V_i^3 A_i$, with power coefficient $C_{P,i}' = a \ctihat$. The coefficient in the latter linear relation is set to $a = 0.9$ to mitigate overprediction of turbine power on present-day simulation grids as discussed in Ref. \cite{munters2017optimal}. Both the thrust coefficient setpoint controls $\cti$ and the yaw rate controls $\omega_i$ are subjected to technical box constraints in Equations \eqref{eq:boxct_constraint} - \eqref{eq:boxomega_constraint}. As discussed in Section \ref{sec:case_definition} and Table \ref{tab:case_definition}, the control cases considered in this manuscript are differentiated based on the parameter values of these box constraints. The system states are grouped in the vector $\bs{q} = [\utilde(\bs{x},t),~\ptilde(\bs{x},t),~\ctnhat{1}(t), \dots, \ctnhat{N_t}(t),~ \theta_1(t), \dots, \theta_{N_t}(t)] = [\utilde(\bs{x},t),~\ptilde(\bs{x},t),~\widehat{\bs{C}}_T'(t) ,~ \bs{\theta}(t)]$ and the controls, consisting of the individual unsteady thrust coefficient setpoints and yaw rates of every turbine, are denoted as $\bs{\varphi} = [\bs{C}_T'(t),~\bs{\omega}(t)]$.%[\ctn{1},\dots,\ctn{N_t},~\omega_1,\dots,\omega_{N_t}]$.

@@

\subsection{Simulation setup}\label{sec:num_setup}
\noindent The simulation domain has a \revision{total dimension of $L_x \times L_y \times L_z = 4.8 \times 2.4 \times 1 $ km$^3$, with $L_x, L_y$, and $L_z$ the streamwise, transversal, and vertical extent of the domain respectively. This domain is discretized on a grid of $N_x \times N_y \times N_z = 256 \times 128 \times 128$ gridpoints.} The final $10\%$ of the streamwise extent of the domain is used as a fringe region to impose inflow conditions. All cases apply periodic boundary conditions on transversal domain boundaries.

Two different sets of simulations are considered. The first set is subjected to a \revision{constant} and uniform inflow with $U_\infty = 8$ m s$^{-1}$. In this case, the top and bottom boundaries are treated with a symmetry condition, \revision{and the vertical location of the turbine hubs $z_h$ is placed at the vertical mid-plane of the domain, i.e. $z_h/L_z = 0.5$,} in order to minimize any influence of vertical domain boundaries. The second set applies unsteady turbulent inflow conditions, which are generated in a pressure-driven precursor simulation of an unperturbed turbulent boundary layer with shifted periodic boundary conditions \cite{munters2016shifted}, driven by a pressure gradient forcing of $\partial_x p_\infty/\rho = 2.5 \times 10^{-4}$ m s$^{-2}$. This results in a freestream wind speed at hub height fluctuating slightly above 8 m s$^{-1}$. In this second set of simulations, the top boundary is treated with a stress-free condition, and a high-Reynolds number wall model is used at the bottom. The turbines are placed at a height of $z_h/L_z = 0.1$ from the wall. The wall model adds a wall stress consistent with a rough-wall logarithmic boundary layer profile with roughness length $z_0/L_z = 10^{-4}$, resulting in a inflow turbulence intensity of about 8\% at hub height.

\revision{Within the context of the receding-horizon methodology introduced in Section~\ref{sec:meth}, the optimization is performed over discrete time windows with optimization horizon $T$ and flow advancement time $T_A$. Ideally, one would take a very large optimization horizon $T$, and a small flow advancement time $T_A$ before recomputing the optimal control problem. However, in practice, $T$ is limited due to the inherent chaotic nature of turbulent flows that complicates control over long time horizons, and $T_A$ should not be taken too small to avoid excessive computational costs. In the current work, we use 6 consecutive windows, each with an optimization horizon $T=300$~s and a flow advancement time of $T_A = T/2= 150$~s, resulting in a total control time of 900~s. Note that, given a wind-farm length of $3 \times 6D = 1 800$ m and a free-stream velocity $U_\infty = 8$ m s$^{-1}$, the wind-farm flowthrough time is approximately 225 s. Hence, the total control time comprises about 4 flowthroughs, and the optimization horizon within each window covers about 1.33 flowthrough times. Because of the latter, each turbine row has the opportunity to take into account interaction with every downstream row, and the sensitivity of power gains to the optimization horizon $T$ is expected to be limited.} All time-averaged results shown below are based on statistics gathered after 300~s, so that wake propagation time lag at startup is excluded. To limit computational efforts, we perform a fixed amount of 80 L--BFGS--B iterations in each window, corresponding to approximately 160 PDE simulations (forward or adjoint). Although formal convergence is not achieved for any of the cases after this amount of iterations, it will be shown later that the relative ordering of the cases based on power extraction is clear. Computational costs amount to approximately 35~000 corehours per optimization case for all 6 time windows combined on 112 Intel Broadwell architecture cores connected by an EDR Infiniband network. The initial guess for the optimal controls within each window is a steady thrust coefficient setpoint $C_T'=2$  and yaw rate $\omega = 0^\circ$ s$^{-1}$ for all turbines. \revision{The sensitivity of attainable power gains to $T/T_A$, initial control guesses, and optimizer settings has been shown to be limited in a prior study \cite{munters2017optimal}.} Simulation and optimization parameters are summarized below in Table~\ref{tab:case_definition}.

Simulations are initialized as follows: for the turbulent inflow case, a fully-developed high-Reynolds number turbulent boundary layer is generated in the precursor inflow simulation by simulating a perturbed logarithmic flow until a statistically-stationary state is achieved. For both the turbulent and the uniform inflow cases the main wind-farm simulation is then advanced in time with the proper inflow conditions for approximately five wind-farm flow-through times, after which the effects of wind-farm start-up transients have subsided. These statistically stationary wind-farm flows then serve as initial conditions for the control cases defined below. Planviews at hub height of these initial conditions are shown in Figure~\ref{fig:initial_conditions_flow_yaw}. It can be seen from the Figure~\ref{fig:initial_conditions_flow_yaw}a that, given laminar inflow conditions, turbine wakes are very deep and remain stable up until row 3, after which meandering and vortex shedding starts to occur. These flow conditions can be regarded as somewhat artificial since, in the atmospheric boundary layer, background turbulence typically causes instant wake transition behind the turbine rotor. \revision{Furthermore, it is documented in literature that numerical details such as grid resolution, subgrid-scale modeling, and turbine models can have a significant impact on simulation results , whereas in turbulent inflow this is much less the case  \cite{martinezwake,sarlakwe,troldborgtorque}. In practice, this tends to lead to artificially stabilized wakes with very low farm efficiencies (see, e.g., Refs \cite{sarlaktorque, ciriacc}).} Nevertheless, this setup allows us to investigate control dynamics more easily, as we can rule out any reaction of the optimization to the background variability. Furthermore, suppressed wake meandering and reduced turbulence levels have been known to occur in wind farms submerged in stably stratified atmospheric boundary layers \cite{larsen2009dependence, machefaux2016experimental}, hence rendering the current case somewhat relevant for such situations. Figure~\ref{fig:initial_conditions_flow_yaw}b shows that, for the turbulent inflow case, the background turbulence causes wake instabilities instantly, leading to enhanced wake recovery and much more complex unsteady wake interactions between turbines.

\begin{figure}
	\includegraphics[width=\textwidth]{figure3_revised}
	\caption{Snapshots of initial conditions for optimal control cases. \emph{a)} Uniform and \revision{constant} inflow. \emph{b)} Turbulent inflow. Colors represent instantaneous streamwise velocity $\widetilde{u}_x$ in m s$^{-1}$. The black lines represent wind turbine locations. The black dashed line indicates the start of the fringe region. R1 -- R4 and C1 -- C4 indicate the numbering convention used for turbine rows and columns respectively. \label{fig:initial_conditions_flow_yaw}}
\end{figure}

@@

Similar to earlier studies \cite{munters2017optimal, munters2016effect}, also two induction control cases are considered, in which turbines can adapt their thrust coefficient $C_T'$ with a characteristic timescale $\tau = 5$ s: an underinductive case where turbines are restricted to lowering $C_T'$ below the Betz-optimal value if desired (I2, with $\ctmax = 2$), and an overinductive case in which also higher $C_T'$ values resulting in increased thrust forces are allowed (I3, with $\ctmax = 3$). Finally, for both the latter cases, also the option of combining dynamic induction control with the yaw control of case Y is considered, defining the yaw--induction cases I2Y and I3Y respectively. The cases investigated in this study are summarized in Table~\ref{tab:case_definition}.

@@

Further results show that, for the current setup \revision{with a relatively small aligned wind farm}, all yaw-enabled cases consistently outperform cases based solely on axial induction control. However, the gap between both is significantly larger for uniform inflow than for turbulent inflow. The main reason for this is that the yawing cases perform better in the absence of inflow turbulence due to the fact that a \revision{constant incoming} flow allows wakes to be redirected more easily.
